---
title: "From Norms to Neuropsychopathology: Exploring Neuroanatomical & Neuropsychiatric Variation in Alzheimer’s Disease Through Normative Modeling"
subtitle: "This study extends analyses conducted by <a href='https://www.neurology.org/doi/10.1212/WNL.0000000000207298'>Verdi et al. (2023)</a>, who utilized normative modeling techniques to delineate neuroanatomical heterogeneity in Alzheimer's disease. We employ a similar methodological framework to supplement these insights by integrating both structural MRI data and neuropsychiatric symptom profiles. This integration allowed us to explore more deeply the neuroanatomical and neuropsychiatric underpinnings of Alzheimer's disease across different phenotypes within our ADNI subset. This study was conducted by the <a href='https://www.mcleanmri.org/ansl.html'>Applied Neuroimaging Statistics Ressearch Laboratory</a> at McLean Hospital & Harvard Medical School."
format:
  lumo-html: 
    logo: "Brain Logo.png"
    github-repo: "https://www.github.com/adrian-a-medina/ADNI_NormativeModeling"
    primary-color: "#a31f37"
    self-contained: true
    is-particlejs-enabled: true
author: Adrian Medina
date: last-modified
---

# Project Overview

[**Analytic Background**]{.underline}**:\
**Our study data were a subset derived from the Alzheimer's Disease Neuroimaging Initiative (ADNI) ADNI3 wave data bank. This was due to the harmonization of scanner sequence protocols across data collection sites that began during this wave. We only included subjects that had both structural MRI and PET (amyloid & tau) data collected, followed by QA of these data. This analysis leveraged a normative model developed by the Predictive Clinical Neuroscience Group at the Donders Institute and Radboud UMC, which aims to predict and stratify brain disorders on the basis of neuroimaging data. Specifically, we used 'HBR_lifespan_36K_79sites', which makes use of the Hierarchical Bayesian Regression algorithm trained on 37,128 subjects from 79 different collection sites, across the human lifespan.

-   Please refer to the Group's [Normative Modeling Graphical User Interface (GUI)](https://pcnportal.dccn.nl/) for more information. For reference to the template Python code used to calculate the deviation scores, please look at the Group's [Braincharts GUI](https://pcntoolkit.readthedocs.io/en/latest/pages/apply_normative_models.html).

[**Considerations**]{.underline}:

-   Ideally, both adaptation and testing sets would be balanced by age, sex, and site (covariates) following something like a 60/40 or 70/30 split of healthy controls. **However**, given our limited sample size, we decided to keep all of our healthy control and patient data **isolated**.
    -   *In our analysis*: The **adaptation set** is used to calibrate for site (**only** [healthy controls]{.underline}) while the **testing set** is used **exclusively** for [patient-phenotyped data]{.underline}.
        -   Healthy control phenotypes include 'A-C-' (amyloid *negative*, cognitive impairment *negative*) & 'A+C-' (amyloid *positive*, cognitive impairment *negative*).
        -   Patient-phenotypes include 'A+C+' (amyloid *positive*, cognitive impairment *positive*) & 'A-C+' (amyloid *negative*, cognitive impairment *positive*).
-   As a consequence of simultaneous conditions, a smaller sample size & larger site numbers (59 sites), our group elected to utilize the MRI manufacturer (3 total) of the subject's imaging data to act as a pseudo 'site' variable thus giving more power to viably calibrate for potential 'site' influences.
    -   '1' = 'GE'
    -   '2' = 'Philips'
    -   '3' = 'Siemens'

# Code Workflow

1.  [Data Frame Initialization](#data-frame-initialization): Install/load packages needed to run any/all chunks in this code file, and load/merge dataset(s) with imaging &/or non-imaging variables of interest. The finalized, merged file is what will be used to create adaptation-testing sets in the next step.
    -   *Note:* Some manual edits may be needed to match your imaging variables to the column headers listed in the normative model CSV template (listed in the [Normative Modeling GUI](https://pcnportal.dccn.nl/)).
2.  [Covariate Distributions, Data Splitting, & Balance Verification](#covariate-distributions-data-splitting-balance-verification): Visualize distributions of the covariates of interest prior to splitting your dataset. Produce the adaptation and testing datasets needed to compute deviation scores of subject neuroimaging (structural MRI) data. Visualize the proportions of covariates of interest following the adaptation-testing split to verify balance between the two sets.
    -   *Note:* Upon completion of this step, you are then able to run your sets through the normative model deviation scores computation via either the [Normative Modeling GUI](https://pcnportal.dccn.nl/) or [Braincharts GUI](https://pcntoolkit.readthedocs.io/en/latest/pages/apply_normative_models.html) (if you want to recreate the Python code yourself).
3.  [Analytic Data Preparation](#analytic-data-preparation): Preparing analytical data for both group comparisons of cortical thickness and broader statistical analyses. Loading and cleaning study data from its CSV file, which includes renaming columns for clarity, re-coding categorical variables such as phenotypes and sites, and setting appropriate factor levels. For analyses excluding group cortical thickness comparisons, the procedure extends to managing deviation scores, generating dummy variables, eliminating unnecessary columns, and ensuring the data's integrity for modeling. Additionally, the Destrieux atlas data is loaded from the `ggseg` package, with preliminary visualizations executed to confirm the structure and detail of the data. This includes plotting atlas dimensions and displaying labeled regions to verify data accuracy. These meticulous preparations ensure the data is aptly formatted and enhanced, ready for subsequent in-depth statistical analysis and visualization tasks.
4.  [Demographic Descriptives and Statistical Analysis of Cortical Thickness](#demographic-descriptives-and-statistical-analysis-of-cortical-thickness): Perform statistical analyses to understand the demographic distribution and cortical thickness variations across different phenotype groups. Calculate the counts for phenotypes, sex, and site categories and provides descriptive statistics for mean cortical thickness and age by phenotype group. Visualizations such as violin plots and bar charts are generated to display the distribution of age and sex across phenotype groups. Modeling cortical thickness variations, fitting linear models to predict mean cortical thickness based solely on phenotype groups and then incorporating additional covariates like age and sex. Detail the model outcomes with confidence intervals, performs Tukey HSD tests to scrutinize mean differences across groups, and visualizes these differences and their statistical significance. Visualize adjusted models' predictions to provide a comprehensive understanding of cortical thickness influences.
5.  [Regional Analyses: Across-Groups & Pairwise Comparisons](#regional-analyses-across-groups-pairwise-comparisons): Apply the suffix '\_rois' to all ROI FreeSurfer measures in your data frame, perform ANOVA analyses to extract p-values for each ROI across phenotype groups, and run FDR corrections. Further, derive F-stat values, merge and organize results for visualization, and display a heatmap of the significant F-stat values, annotated with significance levels. Remove the '\_rois' suffix from your ROI list and ensure that ROI names in your dataset match the Destrieux atlas nomenclature by extensively modifying and standardizing ROI names for both left and right hemispheres, then verifying and assigning corrected names in your data frame. Visualize FDR-corrected p-values using `ggseg` for both left and right hemispheres, employing color gradients to represent significance levels, and save the combined plots as a PDF. Prepare data subsets for pairwise t-tests between different phenotype groups and perform these tests to extract statistical measures such as t-values, p-values, and FDR corrections, then visualize significant results using a lollipop plot to highlight the differences in brain regions across comparisons. Transform ROI names in your dataset to match the Destrieux atlas nomenclature by applying a function that renames based on hemisphere specifications and then ensures consistency with atlas data, using pattern replacements and validation against the atlas's region list. Visualize and compare the significant FDR-corrected p-values across different phenotype groups (Control vs. Alzheimer's, Control vs. MCI, and MCI vs. Alzheimer's) for both hemispheres using `ggseg` maps, adjust the color gradients to highlight significance levels, and then compile all plots into a PDF for detailed analysis.
6.  [Session Information](#session-information): To enhance **reproducibility**, details about the **working environment** used for these analyses can be found below.

# Data Frame Initialization {#data-frame-initialization}

```{r Preliminary Setup, message=FALSE, warning=FALSE}
# Set CRAN repository for consistent package installation, ggseg for neuroimaging visualizations
options(repos = c(
  ggseg = 'https://ggseg.r-universe.dev',
  CRAN = 'https://cloud.r-project.org'))

# Install and load the pacman package for efficient package management
if (!require(pacman)) install.packages("pacman")
library(pacman)

# Use p_load function to install (if necessary) and load packages
p_load(dplyr, knitr, kableExtra, psych, tidyr, readr, stringr, matrixStats, ggseg, ggseg3d, 
       ggsegExtra, ggsegDesterieux, cowplot, data.table, e1071, ggplot2, plot.matrix, proxy, 
       RPMG, broom, gridExtra, patchwork, caret, tidyverse, fastDummies, sjPlot, ggbeeswarm, 
       lavaan, caTools, bitops, effects, ggeffects, reshape2, ggpubr)

# Specify the 'base_path' where you can find your data files, ASSUMING they're in the same directory, & set it as WD
base_path <- "~/GitHub/ADNI_NormativeModeling/data_files"
setwd(base_path)

# Read in foundational datasets
NIDP_DX_Dem_NP <- read_csv("NIDP_DX_Dem_NP_MRI.csv") # Data set with clinical/non-imaging variables and MRI manufacturer information per subject assuming separate from FreeSurfer dataset
ADNI_freesurfer <- read_csv("ADNI_lh_rh_thickness_subcort_volumes.csv") # Data set with FreeSurfer brain thickness values

# Merging datasets on 'Subject_ID' with only ID values present in NIDP_DX_Dem_NP
ADNI_274_Merged <- merge(NIDP_DX_Dem_NP, ADNI_freesurfer, by = "Subject_ID", all.x = TRUE)

# Save the newly merged dataset as a CSV file
write_csv(ADNI_274_Merged, "ADNI_274_Merged.csv")

### Manual edits were done to this spreadsheet to only leave FreeSurfer values and balancing variables for data split
### FreeSurfer variables were edited to match PCN template, see GUI for details
### Covariates: 'age', 'sex', 'site'
# Rename finalized, merged file to 'ADNI_274_Merged_Final.csv'
ADNI_274_Merged_Final <- read_csv("ADNI_274_Merged_Final.csv") # This should be the spreadsheet that will be split into adaptation and testing sets for normative modeling

# Read in computed deviation scores (the file you get back after uploading adaptation and testing sets)
ADNI_deviation_scores <- read_csv("ADNI_deviation_scores_BLR.csv") # This dataset contains raw subject FreeSurfer volumetric measures and their respective deviation scores for your TESTING set used in the Normative Modeling

# Read in PDC deviation data sets
PDC_HBR <- read_csv("PDC_HBR.csv")
PDC_BLR <- read_csv("PDC_BLR.csv")

```

# Covariate Distributions, Data Splitting, & Balance Verification {#covariate-distributions-data-splitting-balance-verification}

```{r phenotype-Site Cross Tabulation Visualizations, message=FALSE, warning=FALSE}
# Create a bar plot for MRI manufacturers with counts displayed
ggplot(NIDP_DX_Dem_NP, aes(x = MRI_MANU)) +
  geom_bar(stat = "count", fill = "skyblue", color = "black") +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
  labs(title = "Count of MRI Manufacturers", x = "MRI Manufacturer", y = "Count") +
  theme_minimal()

# Create separate bar plots for each manufacturer
ggplot(NIDP_DX_Dem_NP, aes(x = MRI_MAKE, fill = MRI_MANU)) +
  geom_bar(position = "dodge") +
  geom_text(aes(label = ..count..), stat = "count", position = position_dodge(width = 0.9), vjust = -0.5) +
  facet_wrap(~MRI_MANU, scales = "free_x") +
  labs(title = "Counts of Each MRI Make Grouped by Manufacturer", x = "MRI Make", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better visibility

# Counting subjects by BioMarkers within each site
bio_markers_site_counts <- ADNI_274_Merged_Final %>%
  group_by(site, BioMarkers) %>%
  dplyr::summarise(Count = n(), .groups = 'drop')

# Plot with Site on x-axis
plot_site_x <- ggplot(bio_markers_site_counts, aes(x = site, y = Count, fill = BioMarkers)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  geom_text(aes(label = Count), vjust = -0.5, position = position_dodge(width = 0.9)) +
  labs(title = "Subject Counts by BioMarkers Across Sites",
       x = "Site",
       y = "Number of Subjects",
       fill = "BioMarkers") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

# Plot with BioMarkers on x-axis
plot_biomarkers_x <- ggplot(bio_markers_site_counts, aes(x = BioMarkers, y = Count, fill = factor(site))) +
  geom_bar(stat = "identity", position = position_dodge()) +  # Use position_dodge() for proper grouping
  geom_text(aes(label = Count), vjust = -0.5, position = position_dodge(width = 0.9)) +  # Adjust text positioning
  labs(title = "Subject Counts Across BioMarkers by Site",
       x = "BioMarkers",
       y = "Number of Subjects",
       fill = "Site") +
  scale_fill_brewer(palette = "Set1") +  # Using a qualitative palette for distinct sites
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))  # Improve readability of x-axis labels


# Optionally, print both plots to view them in the R environment
print(plot_site_x)
print(plot_biomarkers_x)

```

```{r Re-Creating Testing & Adaptation Datasets with Updated Site, message=FALSE, warning=FALSE}
# Separate the data based on 'BioMarkers'
adaptation_data <- ADNI_274_Merged_Final %>%
  filter(BioMarkers %in% c("A-C-", "A+C-"))

testing_data <- ADNI_274_Merged_Final %>%
  filter(BioMarkers %in% c("A-C+", "A+C+"))

# Save the adaptation and testing datasets as CSV files, un-comment when v
# write_csv(adaptation_data, "ADNI_175_Adaptation.csv")
# write_csv(testing_data, "ADNI_99_Testing.csv")

# After verifying that the grouping variable and covariates are balanced as desired (i.e., see the next chunk), you can now run them through the normative model computation via the GUI or recreate their code yourself!

```

```{r Creating Plots to Visualize Dataset Balance Across Split, message=FALSE, warning=FALSE}
# Plotting the distribution of 'BioMarkers' in both datasets
biomarker_distribution <- bind_rows(
  mutate(adaptation_data, dataset = "Adaptation"),
  mutate(testing_data, dataset = "Testing")
) %>%
  group_by(dataset, BioMarkers) %>%
  dplyr::summarise(Count = n(), .groups = 'drop') %>%
  group_by(dataset) %>%
  mutate(Percentage = (Count / sum(Count)) * 100)

ggplot(biomarker_distribution, aes(x = BioMarkers, y = Percentage, fill = dataset)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(title = "Distribution of Biomarker", x = "Biomarker", y = "Percentage (%)") +
  theme_minimal()

# Plotting the distribution of 'site' in both datasets
site_distribution <- bind_rows(
  mutate(adaptation_data, dataset = "Adaptation"),
  mutate(testing_data, dataset = "Testing")
) %>%
  group_by(dataset, site) %>%
  dplyr::summarise(Count = n(), .groups = 'drop') %>%
  group_by(dataset) %>%
  mutate(Percentage = (Count / sum(Count)) * 100)

ggplot(site_distribution, aes(x = site, y = Percentage, fill = dataset)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(title = "Distribution of Site", x = "Site", y = "Percentage (%)") +
  theme_minimal()

# Plotting the distribution of 'sex' in both datasets
sex_distribution <- bind_rows(
  mutate(adaptation_data, dataset = "Adaptation"),
  mutate(testing_data, dataset = "Testing")
) %>%
  group_by(dataset, sex) %>%
  dplyr::summarise(Count = n(), .groups = 'drop') %>%
  group_by(dataset) %>%
  mutate(Percentage = (Count / sum(Count)) * 100)

ggplot(sex_distribution, aes(x = sex, y = Percentage, fill = dataset)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(title = "Distribution of Sex", x = "Sex", y = "Percentage (%)") +
  theme_minimal()

# Calculating and comparing the average age
age_summary <- bind_rows(
  mutate(adaptation_data, dataset = "Adaptation"),
  mutate(testing_data, dataset = "Testing")
) %>%
  group_by(dataset) %>%
  dplyr::summarise(Average_Age = mean(age, na.rm = TRUE), .groups = 'drop')

print(age_summary)

```

# Analytic Data Preparation {#analytic-data-preparation}

```{r Preparing the Destrieux Atlas ROI Variables, message=FALSE, warning=FALSE}
# Assign atlas data to data frames in local environment
desterieux_dims <- desterieux$data
desterieux <- desterieux  # Note: the ggseg package adds an extra 'e' in the spelling of the name Destrieux, so it's NOT a typo

# Plot simple atlas features to test data frame with dimension data
plot(desterieux_dims) +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 7)) +
  guides(fill = guide_legend(ncol = 3))

# Plot atlas ROIs with labels to test data frame with complete vectors
ggplot() +
  geom_brain(atlas = desterieux)

```

```{r Recoding & Reorganizing Analytic Variables for ONLY Group Cortical Thickness Comparisons, message=FALSE, warning=FALSE}
# Rename data frame for efficient code modeling, using dataset with volumetric FreeSurfer measures for ALL subjects
df_merged <- ADNI_274_Merged_Final # to be used ONLY for group cortical thickness comparisons

# Rename phenotype variable
df_merged <- dplyr::rename(df_merged, phenotype = "BioMarkers")

# Specifying the order of the first few columns for easier viewing
desired_order <- c("Subject_ID", "age", "sex", "site", "phenotype")

# Append the remaining column names that are not specified in desired_order
remaining_cols <- setdiff(names(df_merged), desired_order)

# Combine the specified order with the remaining columns
new_order <- c(desired_order, remaining_cols)

# Reorder the columns in df according to new_order
df_merged <- df_merged[, new_order]

# Recode phenotype variable
df_merged$phenotype <- recode(df_merged$phenotype,
                              "A-C-" = "HC",
                              "A+C-" = "HC",
                              "A-C+" = "MCI",
                              "A+C+" = "AD")
df_merged$phenotype <- as.factor(df_merged$phenotype)
df_merged$phenotype <- factor(df_merged$phenotype, levels = c("HC", "MCI", "AD")) # explicitly set the reference level

# Recode site variable
df_merged$site <- as.factor(df_merged$site)
df_merged$site <- gsub("1", "GE", df_merged$site)
df_merged$site <- gsub("2", "Philips", df_merged$site)
df_merged$site <- gsub("3", "Siemens", df_merged$site)

# Recode sex variable, 0 = females 1 = males
df_merged$sex <- factor(df_merged$sex)
df_merged$sex <- gsub("0", "Female", df_merged$sex)
df_merged$sex <- gsub("1", "Male", df_merged$sex)

```

# Demographic Descriptives and Statistical Analysis of Cortical Thickness {#demographic-descriptives-and-statistical-analysis-of-cortical-thickness}

```{r Descriptive Statistics by Variable, message=FALSE, warning=FALSE}
# Count of each variable of interest
table(df_merged$phenotype)
table(df_merged$sex)
table(df_merged$site)

# Cross-Tabulation of average thickness and phenotype group
describeBy(df_merged$Mean_Thickness, df_merged$phenotype)
d<-(describeBy(df_merged$Mean_Thickness, df_merged$phenotype))

# Cross-Tabulation of age and phenotype group
describeBy(df_merged$age, df_merged$phenotype)
mean(df_merged$age)
sd(df_merged$age)

# Generate the violin plot of age stratified by phenotype categories
ggplot(df_merged, aes(x = phenotype, y = age, fill = phenotype)) +
  geom_violin() +
  labs(title = "Distribution of Age Stratified by Phenotype Groups",
       x = "Phenotype Group", y = "Age") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))  # Improve readability of x-axis labels

# Cross-Tabulation of sex & phenotype, visualization
ggplot(df_merged, aes(x = phenotype, fill = sex)) +
  geom_bar(position = "dodge") +
  labs(title = "Distribution of Sex Across Phenotype Groups",
       x = "Phenotype Group",
       y = "Count",
       fill = "Sex") +
  theme_minimal()

# Create the summary Mean_Thickness data frame
filtered_MT_summary <- df_merged %>%
  group_by(phenotype) %>%
  dplyr::summarise(
    score_mean = mean(Mean_Thickness, na.rm = TRUE),
    n = n(), # Sample size for each group
    sem = sd(Mean_Thickness, na.rm = TRUE) / sqrt(n()),
    .groups = 'drop'
  )

# Create the rain cloud plot with separate components
ggplot(df_merged, aes(x = phenotype, y = Mean_Thickness, fill = phenotype, color = phenotype)) +
  PupillometryR::geom_flat_violin(adjust = 1.5, trim = FALSE, alpha = 0.5, position = position_nudge(x = 0.2, y = 0), colour = NA) +
  geom_point(aes(x = as.numeric(phenotype)-0.25, y = Mean_Thickness, colour = phenotype), position = position_jitter(width = .05), size = .5, shape = 20) +
  geom_boxplot(outlier.shape = NA, alpha = 0.5, width = 0.25, position = position_dodge(width = 0.3), colour = "black") +
  geom_point(data = filtered_MT_summary, aes(x = factor(phenotype), y = score_mean), shape = 18, position = position_nudge(x = 0.2)) +
  geom_errorbar(data = filtered_MT_summary, aes(x = factor(phenotype), y = score_mean, ymin = score_mean - sem, ymax = score_mean + sem), width = 0.05, position = position_nudge(x = 0.2)) +
  labs(
    title = "Distribution of Mean Cortical Thickness Stratified by Phenotype Groups", 
    y = "Score", 
    x = "Phenotype", 
    fill = "Phenotype Group",  # Legend title for fill
    color = "Phenotype Group"  # Legend title for color
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    legend.title = element_text(face = "bold")  # Make legend title bold
  )

```

```{r Cortical Thickness Modeling, message=FALSE, warning=FALSE, results='asis'}
# Fit a linear model of Mean_Thickness as a function of phenotype and store it in s_2
s_2 <- lm(Mean_Thickness ~ phenotype, data = df_merged)

# Create a table for the linear model with confidence intervals
tab_model(s_2, title = "Linear Model: Average Cortical Thickness as a Function of Phenotype")

# Perform Tukey's Honest Significant Difference test and store results
tukey_results_s2 <- TukeyHSD(aov(Mean_Thickness ~ phenotype, data = df_merged))
print(tukey_results_s2)
tukey_data_s2 <- as.data.frame(tukey_results_s2$phenotype)
tukey_data_s2$Comparison <- rownames(tukey_data_s2)
tukey_data_s2$significant <- ifelse(tukey_data_s2$`p adj` < 0.05, "Significant", "Not Significant")

# Creating the plot of the Tukey HSD test with significance indication
ggplot(tukey_data_s2, aes(y = Comparison, xmin = lwr, xmax = upr, x = diff)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  geom_errorbarh(aes(height = 0.2, color = significant)) +
  geom_point(aes(x = diff, color = significant), size = 2) +
  scale_color_manual(values = c("Significant" = "red", "Not Significant" = "blue")) +
  labs(title = "Tukey HSD Test Results for Mean Cortical Thickness by Phenotype",
       x = "Differences in Mean Levels of Phenotype",
       y = "Comparison",
       color = "P-value Significance") +
  theme_minimal()

# Fit a linear model of Mean_Thickness as a function of phenotype with added covariates
s_1<-lm(Mean_Thickness ~ phenotype + age + sex, data = df_merged)

# Create a table for the linear model including covariates with confidence intervals
tab_model(s_1, title = "Enhanced Linear Model: Average Cortical Thickness as a Function of Phenotype, Adjusted for Age and Sex")

# Plot multiple regression model
ggpredict(s_1, c(terms = "age", "phenotype", "sex")) |>
  plot() +
  labs(title = "Average Cortical Thickness as a Function of Phenotype, Adjusted for Age and Sex",
       x = "Age",
       y = "Mean Cortical Thickness (mm)",
       caption = "Note: 'phenotype' factors, 'HC' = Healthy Control; 'MCI' = Mild Cognitive Impairment; & 'AD' = Alzheimer's Disease.")

```

# Regional Analyses: Across-Groups & Pairwise Comparisons {#regional-analyses-across-groups-pairwise-comparisons}

```{r Across-Groups Comparisons, message=FALSE, warning=FALSE}
# Apply suffix to all ROI FreeSurfer measures
df.rois <- df_merged %>% rename_at(vars((6:153)), ~ paste0(., '_rois'))

# Run ANOVA analyses for each ROI across phenotype group and extract just the p-values
df.stats <- as.data.frame(sapply(X = df.rois[,grep("_rois", names(df.rois),value = T)], FUN = function(x) summary(aov(x ~ df.rois$phenotype))[[1]][["Pr(>F)"]][1]))

# Rename columns in ANOVA output data frame
names(df.stats) <- "p_value"
setDT(df.stats, keep.rownames = "ROI")

# Verify that changes were made via the first 20 ROIs
head(df.stats, 20)

# Run False Discovery Rate (FDR) corrections 
df.stats <- cbind(df.stats, p.adjust(df.stats$p_value), method = "fdr")

# Rename column of FDR-corrected p-values
names(df.stats)[3] <- "FDR.pvalue"

# Verify that changes were made
head(df.stats, 20)

### Repeat steps to now obtain the F-stat values
# Run ANOVA analyses for each ROI across phenotype group and extract just the F-stat values
df.f_stats <- as.data.frame(sapply(X = df.rois[,grep("_rois", names(df.rois),value = T)], FUN = function(x) summary(aov(x ~ df.rois$phenotype))[[1]][["F value"]][1]))

# Rename columns in ANOVA output data frame
names(df.f_stats) <- "f_stat"
setDT(df.f_stats, keep.rownames = "ROI")

# Verify that changes were made via the first 20 ROIs
head(df.f_stats, 20)

# List the ROIs that are significant after FDR correction
df.stats[which(df.stats$FDR.pvalue <0.05),]

# Assign significance levels
df.stats$Significance <- ifelse(df.stats$FDR.pvalue < 0.001, '***',
                               ifelse(df.stats$FDR.pvalue < 0.01, '**',
                               ifelse(df.stats$FDR.pvalue < 0.05, '*', '')))

# Merge data frames for visualization
df.stats_merged <- merge(df.stats, df.f_stats, by = "ROI")

# Melt the data for plotting
df.melted <- melt(df.stats_merged, id.vars = "ROI", variable.name = "Statistic", value.name = "Value")

# Ensure that 'Value' is numeric
df.melted$Value <- as.numeric(as.character(df.melted$Value))

# Ensure df.stats and df.melted are merged to filter and sort
df.melted <- df.melted %>%
  filter(Statistic == "f_stat") %>%
  left_join(df.stats %>% select(ROI, Significance, FDR.pvalue), by = "ROI") %>%
  filter(FDR.pvalue < 0.05) %>%
  arrange(match(Significance, c("", "*", "**", "***")))  # Order by significance levels

# Remove suffix from ROI list
df.melted$ROI <- gsub("_rois", "", df.melted$ROI)

# Define group positions and labels
group_positions <- data.frame(
  start = c(1, 15, 30),  # example start positions
  end = c(14, 29, 51),    # example end positions
  label = c("*", "**", "***")  # example labels for significance
)

# Create the heatmap
heatmap_plot <- ggplot(df.melted, aes(x = ROI, y = Statistic, fill = Value)) +
  geom_tile() +
  scale_fill_gradient(low = "blue", high = "red") +
  labs(title = "Heatmap of ANOVA Statistics Across ROIs", x = "Region of Interest", y = "F-Statistic") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

# Add custom brackets using annotations
for (i in 1:nrow(group_positions)) {
  heatmap_plot <- heatmap_plot +
    annotate("segment", x = group_positions$start[i], xend = group_positions$end[i], y = Inf, yend = Inf, yjust = 1.5, color = "black", size = 0.5) +
    annotate("text", x = (group_positions$start[i] + group_positions$end[i]) / 2, y = Inf, label = group_positions$label[i], vjust = 2, hjust = 0.5)
}

print(heatmap_plot)

```

```{r Convert Across-Groups Data to Destrieux Atlas Format, message=FALSE, warning=FALSE}
# Remove suffix from ROI list
df.stats$ROI <- as.data.frame(gsub("_rois", "", df.stats$ROI))

### Rename ROIs to assign nomenclature consistent with the 'desterieux' atlas package
# Left hemisphere ROIs
left.df.stats <- df.stats %>%
  filter(str_detect(ROI, "L_"))
x <- gsub("L_", "", left.df.stats$ROI)
y <- gsub("G&S_", "G and S ", x)
z <- gsub("G_", "G ", y)
z1 <- gsub("S_", "S ", z)
z2 <- gsub("_", " ", z1)
z3 <- gsub(" bin", "", z2)
z4 <- gsub("\\.", " ", z3)
z5 <- gsub("cingul ", "cingul-", z4)
z6 <- gsub("Mid ", "Mid-", z5)
z7 <- gsub("Post ", "Post-", z6)
z8 <- gsub("inf ", "inf-", z7)
z9 <- gsub("med ", "med-", z8)
z10 <- gsub("sup ", "sup-", z9)
z11 <- gsub("Fis ant ", "Fis-ant-", z10)
z12 <- gsub("precentral ", "precentral-", z11)
z13 <- gsub("Fis pos ", "Fis-pos-", z12)
z14 <- gsub("lg&S", "lg and S", z13)
z15 <- gsub("oc temp", "oc-temp", z14)
z16 <- gsub("sup and transversal", "sup-transversal", z15)
z17 <- gsub("orbital H Shaped", "orbital-H Shaped", z16)
z18 <- gsub("oc sup&transversal", "oc sup and transversal", z17)
z19 <- gsub("prim Jensen", "prim-Jensen", z18)
z20 <- gsub("S oc-temp med&Lingual", "S oc-temp med and Lingual", z19)
z21 <- gsub("lat fusifor", "lat-fusifor", z20)
z22 <- gsub("middle&Lunatus", "middle and Lunatus", z21)
z23 <- gsub("intrapariet&P trans", "intrapariet and P trans", z22)
renamed_ROIs <- gsub("Lat Fis post", "Lat Fis-post", z23)

desterieux_ROIs <- as.data.frame(desterieux_dims %>% filter(hemi == "left"))$region
compare_lists <- cbind(sort(renamed_ROIs), sort(unique(desterieux_ROIs)))
list_matches <- compare_lists[,1] %in% compare_lists[,2]
compare_lists[!list_matches,]

## if no mismatches, than add to data.frame as region
left.df.stats$region <- renamed_ROIs

# Right hemisphere ROIs
right.df.stats <- df.stats %>%
  filter(str_detect(ROI, "R_"))
x <- gsub("R_", "", right.df.stats$ROI)
y <- gsub("G&S_", "G and S ", x)
z <- gsub("G_", "G ", y)
z1 <- gsub("S_", "S ", z)
z2 <- gsub("_", " ", z1)
z3 <- gsub(" bin", "", z2)
z4 <- gsub("\\.", " ", z3)
z5 <- gsub("cingul ", "cingul-", z4)
z6 <- gsub("Mid ", "Mid-", z5)
z7 <- gsub("Post ", "Post-", z6)
z8 <- gsub("inf ", "inf-", z7)
z9 <- gsub("med ", "med-", z8)
z10 <- gsub("sup ", "sup-", z9)
z11 <- gsub("Fis ant ", "Fis-ant-", z10)
z12 <- gsub("precentral ", "precentral-", z11)
z13 <- gsub("Fis pos ", "Fis-pos-", z12)
z14 <- gsub("lg&S", "lg and S", z13)
z15 <- gsub("oc temp", "oc-temp", z14)
z16 <- gsub("sup and transversal", "sup-transversal", z15)
z17 <- gsub("orbital H Shaped", "orbital-H Shaped", z16)
z18 <- gsub("oc sup&transversal", "oc sup and transversal", z17)
z19 <- gsub("prim Jensen", "prim-Jensen", z18)
z20 <- gsub("S oc-temp med&Lingual", "S oc-temp med and Lingual", z19)
z21 <- gsub("lat fusifor", "lat-fusifor", z20)
z22 <- gsub("middle&Lunatus", "middle and Lunatus", z21)
z23 <- gsub("intrapariet&P trans", "intrapariet and P trans", z22)
renamed_ROIs <- gsub("Lat Fis post", "Lat Fis-post", z23)

desterieux_ROIs <- as.data.frame(desterieux_dims %>% filter(hemi == "right"))$region
compare_lists <- cbind(sort(renamed_ROIs), sort(unique(desterieux_ROIs)))
list_matches <- compare_lists[,1] %in% compare_lists[,2]
compare_lists[!list_matches,]

## if no mismatches, than add to data.frame as region
right.df.stats$region <- renamed_ROIs

```

```{r Plot Significant FDR-Corrected p-Values for Across-Groups Data, message=FALSE, warning=FALSE}
# Left hemisphere
left_pvalues <- ggseg(.data=left.df.stats, atlas = desterieux, mapping=aes(fill=FDR.pvalue), hemisphere = "left", colour = "white", size = 0.2) + 
scale_fill_gradientn(limits = c(0,0.05), colours =  rainbow.colors(5))

# Right hemisphere
right_pvalues <- ggseg(.data=right.df.stats, atlas = desterieux, mapping=aes(fill=FDR.pvalue), hemisphere = "right", colour = "white", size = 0.2) +
  scale_fill_gradientn(limits = c(0,0.05), colours = rainbow.colors(5))

# Add titles to individual plots
left_pvalues <- left_pvalues + labs(title = "Between-Group Comparisons")
right_pvalues <- right_pvalues + labs(title = "Between-Group Comparisons")

cowplot::plot_grid(left_pvalues,right_pvalues, nrow = 2, labels = "AUTO")

```

```{r Pairwise Comparisons, message=FALSE, warning=FALSE}
# Data frame preparation
df.CA <- df_merged[grep("HC|AD", df_merged$phenotype), ] # Subset just Controls and Alzheimer's ("CA")
df.CM <- df_merged[grep("HC|MCI", df_merged$phenotype), ] # Subset just Controls and MCI ("CM")
df.MA <- df_merged[grep("MCI|AD", df_merged$phenotype), ] # Subset just MCI and Alzheimer's ("MA")

### Control vs Alzheimer's t-test
# Perform t-test and extract t-stat value, p-value, and parameters
df.CA_stats <- as.data.frame(sapply(df.CA[6:153], function(x) t.test(x ~ df.CA$phenotype)$statistic))
df.CA_stats1 <- as.data.frame(sapply(df.CA[6:153], function(x) t.test(x ~ df.CA$phenotype)$p.value))
df.CA_stats2 <- as.data.frame(sapply(df.CA[6:153], function(x) t.test(x ~ df.CA$phenotype)$parameter))

# Merge data frames and remove the subsequent, intermediate data frames
df.CA_stats <- cbind(df.CA_stats, df.CA_stats1)
df.CA_stats <- cbind(df.CA_stats, df.CA_stats2)
rm(df.CA_stats1, df.CA_stats2)

# Clean up the t-test data frame
df.CA_stats$t_statistic <- df.CA_stats[2] # Rename column
names(df.CA_stats) <- c('statistic', 'p.value', 'parameter')
df.CA_stats[4] <- NULL

# Perform FDR correction on t-test data frame
df.CA_stats <- cbind(df.CA_stats, p.adjust(df.CA_stats$p.value), method = "fdr")
names(df.CA_stats)[4] <- "FDR.pvalue"
df.CA_stats[5] <- NULL

# Clean up the FDR-corrected data frame
df.CA_stats <- tibble::rownames_to_column(df.CA_stats, "ROI") # Create ROI column
df.CA_stats$ROI <- gsub("\\.t$", "", df.CA_stats$ROI)

# List the ROIs that are significant after FDR correction
df.CA_stats_sig <- df.CA_stats[which(df.CA_stats$FDR.pvalue <0.05),]

### Control vs MCI t-test
# Perform t-test and extract t-stat value, p-value, and parameters
df.CM_stats <- as.data.frame(sapply(df.CM[6:153], function(x) t.test(x ~ df.CM$phenotype)$statistic))
df.CM_stats1 <- as.data.frame(sapply(df.CM[6:153], function(x) t.test(x ~ df.CM$phenotype)$p.value))
df.CM_stats2 <- as.data.frame(sapply(df.CM[6:153], function(x) t.test(x ~ df.CM$phenotype)$parameter))

# Merge data frames and remove the subsequent, intermediate data frames
df.CM_stats <- cbind(df.CM_stats, df.CM_stats1)
df.CM_stats <- cbind(df.CM_stats, df.CM_stats2)
rm(df.CM_stats1, df.CM_stats2)

# Clean up the t-test data frame
df.CM_stats$t_statistic <- df.CM_stats[2] # Rename column
names(df.CM_stats) <- c('statistic', 'p.value', 'parameter')
df.CM_stats[4] <- NULL

# Perform FDR correction on t-test data frame
df.CM_stats <- cbind(df.CM_stats, p.adjust(df.CM_stats$p.value), method = "fdr")
names(df.CM_stats)[4] <- "FDR.pvalue"
df.CM_stats[5] <- NULL

# Clean up the FDR-corrected data frame
df.CM_stats <- tibble::rownames_to_column(df.CM_stats, "ROI") # Create ROI column
df.CM_stats$ROI <- gsub("\\.t$", "", df.CM_stats$ROI)

# List the ROIs that are significant after FDR correction
df.CM_stats_sig <- df.CM_stats[which(df.CM_stats$FDR.pvalue <0.05),]

### MCI vs Alzheimer's t-test
# Perform t-test and extract t-stat value, p-value, and parameters
df.MA_stats <- as.data.frame(sapply(df.MA[6:153], function(x) t.test(x ~ df.MA$phenotype)$statistic))
df.MA_stats1 <- as.data.frame(sapply(df.MA[6:153], function(x) t.test(x ~ df.MA$phenotype)$p.value))
df.MA_stats2 <- as.data.frame(sapply(df.MA[6:153], function(x) t.test(x ~ df.MA$phenotype)$parameter))

# Merge data frames and remove the subsequent, intermediate data frames
df.MA_stats <- cbind(df.MA_stats, df.MA_stats1)
df.MA_stats <- cbind(df.MA_stats, df.MA_stats2)
rm(df.MA_stats1, df.MA_stats2)

# Clean up the t-test data frame
df.MA_stats$t_statistic <- df.MA_stats[2] # Rename column
names(df.MA_stats) <- c('statistic', 'p.value', 'parameter')
df.MA_stats[4] <- NULL

# Perform FDR correction on t-test data frame
df.MA_stats <- cbind(df.MA_stats, p.adjust(df.MA_stats$p.value), method = "fdr")
names(df.MA_stats)[4] <- "FDR.pvalue"
df.MA_stats[5] <- NULL

# Clean up the FDR-corrected data frame
df.MA_stats <- tibble::rownames_to_column(df.MA_stats, "ROI") # Create ROI column
df.MA_stats$ROI <- gsub("\\.t$", "", df.MA_stats$ROI)

# List the ROIs that are significant after FDR correction
df.MA_stats_sig <- df.MA_stats[which(df.MA_stats$FDR.pvalue <0.05),]

# Combine the significant results for easy plotting
combined_stats_sig <- bind_rows(
  df.CA_stats_sig %>% mutate(Comparison = "HC vs AD"),
  df.MA_stats_sig %>% mutate(Comparison = "MCI vs AD")
)

# Assign significance levels
combined_stats_sig$Significance <- ifelse(combined_stats_sig$FDR.pvalue < 0.001, '***',
                               ifelse(combined_stats_sig$FDR.pvalue < 0.01, '**',
                               ifelse(combined_stats_sig$FDR.pvalue < 0.05, '*', '')))

# Create the lollipop plot
ggplot(combined_stats_sig, aes(x = reorder(ROI, statistic), y = statistic, color = Comparison)) +
  geom_segment(aes(yend = 0, xend = ROI), size = 1, linetype = "dashed") +
  geom_point(size = 3) +
  labs(
    title = "Significant Differences in Brain Regions Across Comparisons",
    x = "Region of Interest",
    y = "T-Statistic Value",
    color = "Comparison",
    caption = "Note: 'HC' = Healthy Control; 'MCI' = Mild Cognitive Impairment; & 'AD' = Alzheimer's Disease.") +
  theme_minimal() +
  theme(
    legend.position = "top",
    axis.text.x = element_text(angle = 90, hjust = 1, size = 7.5)  # Rotate x labels for better visibility
  )

```

```{r Convert Pairwise Data to Destrieux Atlas Format, message=FALSE, warning=FALSE}
### Rename ROIs to assign nomenclature consistent with the `desterieux` atlas package
# Automated loop version
rename_rois <- function(df, hemi, desterieux_dims) {
  # Filter based on hemisphere
  hemi_prefix <- ifelse(hemi == "left", "L_", "R_")
  df_filtered <- df %>%
    filter(str_detect(ROI, hemi_prefix))
  
  # Perform renaming steps
  x <- gsub(paste0(hemi_prefix), "", df_filtered$ROI)
  patterns <- c("G&S_", "G_", "S_", "_", " bin", "\\.", "cingul ", "Mid ", "Post ", 
                "inf ", "med ", "sup ", "Fis ant ", "precentral ", "Fis pos ", "lg&S", 
                "oc temp", "sup and transversal", "orbital H Shaped", "oc sup&transversal", 
                "prim Jensen", "S oc-temp med&Lingual", "lat fusifor", "middle&Lunatus", 
                "intrapariet&P trans", "Lat Fis post")
  replacements <- c("G and S ", "G ", "S ", " ", "", " ", "cingul-", "Mid-", "Post-", 
                    "inf-", "med-", "sup-", "Fis-ant-", "precentral-", "Fis-pos-", 
                    "lg and S", "oc-temp", "sup-transversal", "orbital-H Shaped", 
                    "oc sup and transversal", "prim-Jensen", "S oc-temp med and Lingual", 
                    "lat-fusifor", "middle and Lunatus", "intrapariet and P trans", 
                    "Lat Fis-post")
  for (i in seq_along(patterns)) {
    x <- gsub(patterns[i], replacements[i], x)
  }
  
  # Match with desterieux ROIs
  desterieux_ROIs <- as.data.frame(desterieux_dims %>% filter(hemi == hemi))$region
  compare_lists <- cbind(sort(x), sort(unique(desterieux_ROIs)))
  list_matches <- compare_lists[,1] %in% compare_lists[,2]
  
  if (any(!list_matches)) {
    warning("There are mismatches in ROI names for the ", hemi, " hemisphere.")
  }
  
  df_filtered$region <- x
  return(df_filtered)
}

# Apply the function to each dataset and hemisphere
left.df.CA_stats <- rename_rois(df.CA_stats, "left", desterieux_dims)
right.df.CA_stats <- rename_rois(df.CA_stats, "right", desterieux_dims)
left.df.CM_stats <- rename_rois(df.CM_stats, "left", desterieux_dims)
right.df.CM_stats <- rename_rois(df.CM_stats, "right", desterieux_dims)
left.df.MA_stats <- rename_rois(df.MA_stats, "left", desterieux_dims)
right.df.MA_stats <- rename_rois(df.MA_stats, "right", desterieux_dims)

```

```{r Plot Significant FDR-Corrected p-Values for Pairwise Data, message=FALSE, warning=FALSE}
### Control vs Alzheimer's
# Left hemisphere
left_CA_pvalues <- ggseg(.data=left.df.CA_stats, atlas = desterieux, mapping=aes(fill=FDR.pvalue), hemisphere = "left", colour = "white", size = 0.2) + 
scale_fill_gradientn(limits = c(0,0.05), colours =  rainbow.colors(5))

# Right hemisphere
right_CA_pvalues <- ggseg(.data=right.df.CA_stats, atlas = desterieux, mapping=aes(fill=FDR.pvalue), hemisphere = "right", colour = "white", size = 0.2) +
  scale_fill_gradientn(limits = c(0,0.05), colours = rainbow.colors(5))

### Control vs MCI
# Left hemisphere
left_CM_pvalues <- ggseg(.data=left.df.CM_stats, atlas = desterieux, mapping=aes(fill=FDR.pvalue), hemisphere = "left", colour = "white", size = 0.2) + 
scale_fill_gradientn(limits = c(0,0.05), colours =  rainbow.colors(5))

# Right hemisphere
right_CM_pvalues <- ggseg(.data=right.df.CM_stats, atlas = desterieux, mapping=aes(fill=FDR.pvalue), hemisphere = "right", colour = "white", size = 0.2) +
  scale_fill_gradientn(limits = c(0,0.05), colours = rainbow.colors(5))

### MCI vs Alzheimer's
# Left hemisphere
left_MA_pvalues <- ggseg(.data=left.df.MA_stats, atlas = desterieux, mapping=aes(fill=FDR.pvalue), hemisphere = "left", colour = "white", size = 0.2) + 
scale_fill_gradientn(limits = c(0,0.05), colours =  rainbow.colors(5))

# Right hemisphere
right_MA_pvalues <- ggseg(.data=right.df.MA_stats, atlas = desterieux, mapping=aes(fill=FDR.pvalue), hemisphere = "right", colour = "white", size = 0.2) +
  scale_fill_gradientn(limits = c(0,0.05), colours = rainbow.colors(5))

# Add titles to individual plots
left_CA_pvalues <- left_CA_pvalues + labs(title = "Healthy Controls vs Alzheimer's Group")
right_CA_pvalues <- right_CA_pvalues + labs(title = "Healthy Controls vs Alzheimer's Group")

left_CM_pvalues <- left_CM_pvalues + labs(title = "Healthy Controls vs MCI Group")
right_CM_pvalues <- right_CM_pvalues + labs(title = "Healthy Controls vs MCI Group")

left_MA_pvalues <- left_MA_pvalues + labs(title = "MCI Group vs Alzheimer's Group")
right_MA_pvalues <- right_MA_pvalues + labs(title = "MCI Group vs Alzheimer's Group")

# Print plots to verify that they were correctly constructed
cowplot::plot_grid(left_CA_pvalues, right_CA_pvalues, nrow = 2)
cowplot::plot_grid(left_CM_pvalues, right_CM_pvalues, nrow = 2)
cowplot::plot_grid(left_MA_pvalues, right_MA_pvalues, nrow = 2)

```

```{r BLR vs HBR Correlation}
# Merge datasets on SubjectID
merged_data <- inner_join(PDC_HBR, PDC_BLR, by = "SubjectID", suffix = c(".HBR", ".BLR"))

# Specify the brain regions for which correlations will be computed for
brain_regions <- c("L_G&S_frontomargin", "L_G&S_occipital_inf", "L_G&S_paracentral", 
                   "L_G&S_subcentral", "L_G&S_transv_frontopol", "L_G&S_cingul-Ant",
                   "L_G&S_cingul-Mid-Ant", "L_G&S_cingul-Mid-Post", "L_G_cingul-Post-dorsal",
                   "L_G_cingul-Post-ventral", "L_G_cuneus", "L_G_front_inf-Opercular", 
                   "L_G_front_inf-Orbital", "L_G_front_inf-Triangul", "L_G_front_middle", 
                   "L_G_front_sup", "L_G_Ins_lg&S_cent_ins", "L_G_insular_short", 
                   "L_G_occipital_middle", "L_G_occipital_sup", "L_G_oc-temp_lat-fusifor",
                   "L_G_oc-temp_med-Lingual", "L_G_oc-temp_med-Parahip", "L_G_orbital", 
                   "L_G_pariet_inf-Angular", "L_G_pariet_inf-Supramar", "L_G_parietal_sup", 
                   "L_G_postcentral", "L_G_precentral", "L_G_precuneus", "L_G_rectus", 
                   "L_G_subcallosal", "L_G_temp_sup-G_T_transv", "L_G_temp_sup-Lateral",
                   "L_G_temp_sup-Plan_polar", "L_G_temp_sup-Plan_tempo", "L_G_temporal_inf",
                   "L_G_temporal_middle", "L_Lat_Fis-ant-Horizont", "L_Lat_Fis-ant-Vertical",
                   "L_Lat_Fis-post", "L_Pole_occipital", "L_Pole_temporal", "L_S_calcarine", 
                   "L_S_central", "L_S_cingul-Marginalis", "L_S_circular_insula_ant",
                   "L_S_circular_insula_inf", "L_S_circular_insula_sup", "L_S_collat_transv_ant", 
                   "L_S_collat_transv_post", "L_S_front_inf", "L_S_front_middle", "L_S_front_sup", 
                   "L_S_interm_prim-Jensen", "L_S_intrapariet&P_trans", "L_S_oc_middle&Lunatus", 
                   "L_S_oc_sup&transversal", "L_S_occipital_ant", "L_S_oc-temp_lat", 
                   "L_S_oc-temp_med&Lingual", "L_S_orbital_lateral", "L_S_orbital_med-olfact", 
                   "L_S_orbital-H_Shaped", "L_S_parieto_occipital", "L_S_pericallosal", 
                   "L_S_postcentral", "L_S_precentral-inf-part", "L_S_precentral-sup-part", 
                   "L_S_suborbital", "L_S_subparietal", "L_S_temporal_inf", "L_S_temporal_sup", 
                   "L_S_temporal_transverse", "R_G&S_frontomargin", "R_G&S_occipital_inf", 
                   "R_G&S_paracentral", "R_G&S_subcentral", "R_G&S_transv_frontopol", 
                   "R_G&S_cingul-Ant", "R_G&S_cingul-Mid-Ant", "R_G&S_cingul-Mid-Post", 
                   "R_G_cingul-Post-dorsal", "R_G_cingul-Post-ventral", "R_G_cuneus", 
                   "R_G_front_inf-Opercular", "R_G_front_inf-Orbital", "R_G_front_inf-Triangul", 
                   "R_G_front_middle", "R_G_front_sup", "R_G_Ins_lg&S_cent_ins", 
                   "R_G_insular_short", "R_G_occipital_middle", "R_G_occipital_sup", 
                   "R_G_oc-temp_lat-fusifor", "R_G_oc-temp_med-Lingual", "R_G_oc-temp_med-Parahip", 
                   "R_G_orbital", "R_G_pariet_inf-Angular", "R_G_pariet_inf-Supramar", 
                   "R_G_parietal_sup", "R_G_postcentral", "R_G_precentral", "R_G_precuneus", 
                   "R_G_rectus", "R_G_subcallosal", "R_G_temp_sup-G_T_transv", "R_G_temp_sup-Lateral",
                   "R_G_temp_sup-Plan_polar", "R_G_temp_sup-Plan_tempo", "R_G_temporal_inf", 
                   "R_G_temporal_middle", "R_Lat_Fis-ant-Horizont", "R_Lat_Fis-ant-Vertical", 
                   "R_Lat_Fis-post", "R_Pole_occipital", "R_Pole_temporal", "R_S_calcarine", 
                   "R_S_central", "R_S_cingul-Marginalis", "R_S_circular_insula_ant", 
                   "R_S_circular_insula_inf", "R_S_circular_insula_sup", "R_S_collat_transv_ant", 
                   "R_S_collat_transv_post", "R_S_front_inf", "R_S_front_middle", "R_S_front_sup", 
                   "R_S_interm_prim-Jensen", "R_S_intrapariet&P_trans", "R_S_oc_middle&Lunatus", 
                   "R_S_oc_sup&transversal", "R_S_occipital_ant", "R_S_oc-temp_lat", 
                   "R_S_oc-temp_med&Lingual", "R_S_orbital_lateral", "R_S_orbital_med-olfact", 
                   "R_S_orbital-H_Shaped", "R_S_parieto_occipital", "R_S_pericallosal", 
                   "R_S_postcentral", "R_S_precentral-inf-part", "R_S_precentral-sup-part", 
                   "R_S_suborbital", "R_S_subparietal", "R_S_temporal_inf", "R_S_temporal_sup", 
                   "R_S_temporal_transverse")

# Compute correlations for each brain region
correlation_results <- list()

for(region in brain_regions) {
    region_hbr <- paste(region, ".HBR", sep = "")
    region_blr <- paste(region, ".BLR", sep = "")
    
    # Ensure both variables are present in the dataset
    if(region_hbr %in% names(merged_data) && region_blr %in% names(merged_data)) {
        # Use cor.test to get both correlation and p-value
        test_result <- cor.test(merged_data[[region_hbr]], merged_data[[region_blr]], 
                                method = "pearson", use = "complete.obs")
        
        # Store results in a list
        correlation_results[[region]] <- list(
            Correlation = test_result$estimate,
            P_Value = test_result$p.value
        )
    }
}

# Convert the list to a data frame for easy viewing
correlation_df <- data.frame(
    Region = names(correlation_results),
    Correlation = sapply(correlation_results, function(x) x$Correlation),
    P_Value = sapply(correlation_results, function(x) x$P_Value),
    row.names = NULL
)

# Print the results
print(correlation_df)

# Convert the data to a suitable format for ggplot
correlation_df$Region <- factor(correlation_df$Region, levels = correlation_df$Region)

# Create a scatter plot for each region showing the correlation values
# Assuming 'merged_data' contains corresponding columns for HBR and BLR scores
p <- ggplot(correlation_df, aes(x = Region, y = Correlation)) +
  geom_point(stat = "identity") +
  geom_smooth(method = "lm", color = "blue") +
  ylim(0.8, 1) +
  labs(title = "Correlation of Deviation Scores Across Brain Regions",
       x = "Brain Region", y = "Correlation Coefficient") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 5.5))

# Print the plot
print(p)

# Adding a new column for significance level based on p-values
correlation_df$Significance <- ifelse(correlation_df$P_Value < 0.001, '***',
                                      ifelse(correlation_df$P_Value < 0.01, '**',
                                      ifelse(correlation_df$P_Value < 0.05, '*', 'ns')))

# Create the lollipop plot
ggplot(correlation_df, aes(x = reorder(Region, Correlation), y = Correlation, color = Significance)) +
    geom_segment(aes(xend = Region, yend = 0.8), size = 0.25) +  # Create lines from 0 to the correlation value
    geom_point(size = 1) +  # Add points for each correlation value
    scale_color_manual(values = c('***' = 'red', '**' = 'orange', '*' = 'blue', 'ns' = 'grey')) +  # Assign colors based on significance
    labs(
        title = "Correlation and Significance of Brain Regions: Hierarchical vs. Linear Bayesian Regression",
        x = "Brain Region",
        y = "Correlation Coefficient",
        color = "Significance Level"
    ) +
    theme_minimal() +
    theme(
        legend.position = "top",  # Move the legend to the top of the plot
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 5.5),  # Improve readability of x-axis labels
        legend.title = element_text(face = "bold")  # Bold the legend title for emphasis
    )

```

# Z-Score Statistics Pipeline

```{r Recoding & Reorganizing Analytic Variables for All Other Analyses, message=FALSE, warning=FALSE}
# Use dummy_cols to create dummy variables for the 'BioMarkers' column
ADNI_deviation_scores <- dummy_cols(ADNI_deviation_scores, select_columns = "BioMarkers", remove_first_dummy = TRUE, remove_selected_columns = TRUE)

# Rename data frame for efficient code modeling, this is the data frame that will be used for all other analyses EXCEPT group cortical thickness comparisons
df <- ADNI_deviation_scores

# Omit the 'index' column from the data frame, unnecessary variable
df <- df[, -which(names(df) == "index")]

# Rename biomarker variable
df <- dplyr::rename(df, phenotype = `BioMarkers_A+C+`)

# Specifying the order of the first few columns for easier viewing
desired_order <- c("Subject_ID", "age", "sex", "site", "phenotype")

# Append the remaining column names that are not specified in desired_order
remaining_cols <- setdiff(names(df), desired_order)

# Combine the specified order with the remaining columns
new_order <- c(desired_order, remaining_cols)

# Reorder the columns in df according to new_order
df <- df[, new_order]

# List of columns to exclude from the ROI list
desired_order <- c("Subject_ID", "age", "sex", "site", "phenotype")

# Extract column names, remove those in desired_order, and remove z-score suffixes
ROI <- names(df)[!names(df) %in% desired_order & !grepl("_Z_predict", names(df))]

# Further clean the ROI names to ensure no duplicates from z-score names
ROI <- unique(gsub("_Z_predict", "", ROI))

# Recode phenotype variable
df$phenotype <- as.factor(df$phenotype)
df$phenotype <- gsub("0", "MCI", df$phenotype)
df$phenotype <- gsub("1", "AD", df$phenotype)
df$phenotype <- factor(df$phenotype, levels = c("MCI", "AD")) # explicitly set the reference level

# Recode site variable
df$site <- as.factor(df$site)
df$site <- gsub("1", "GE", df$site)
df$site <- gsub("2", "Philips", df$site)
df$site <- gsub("3", "Siemens", df$site)

# Recode sex variable, 0= females 1= males
df$sex <- factor(df$sex)
df$sex <- gsub("0", "Female", df$sex)
df$sex <- gsub("1", "Male", df$sex)

```

```{r Binarize Outliers, Create Total Outlier Score Across Total ROIs, and Brief Analyses message=FALSE, warning=FALSE}
# Establish outlier threshold
outlier_threshold <- -1.96 # bottom 2.5%, as used in the Verdi et al., 2023 paper

# Apply threshold to z-score data to create dummy columns
df3 <- as.data.frame(ifelse(df[,6:153] < outlier_threshold,1,0))

# Rename all binarized columns to have the suffix "_bin"
df3 <- df3 %>% rename_all(paste0, "_bin")

# Sum the dummy column values to create a total outlier score
df$total_outlier_score <- rowSums(df3)

# Merge the two data frames
df <- cbind(df, df3)

# Remove df3 from your R environment as it's no longer needed
rm(df3)

# Create the box plot with colored outlines and filled points
ggplot(df, aes(x = phenotype, y = total_outlier_score, color = phenotype, fill = phenotype)) +
  geom_boxplot(outlier.shape = NA, fill = NA, size = 1) +  # Draw box plots with no fill, only outlines
  geom_jitter(width = 0.2, size = 2, shape = 21) +  # Add jittered points with the same color
  labs(x = "Phenotype", y = "Total Outlier Score") +
  theme_minimal() +
  theme(legend.position = "right")

# Calculate descriptives of total outlier score grouped by phenotype group
describeBy(df$total_outlier_score, df$phenotype, IQR = T)

# Basic regression analysis with phenoype as predictor and total outlier score as outcome
summary(lm(total_outlier_score ~ phenotype, data = df))
s1<- lm(total_outlier_score ~ phenotype, data = df)
confint(s1) # Provides CI
# rm(s1)

```

# Session Information {#session-information}

```{r}
sessionInfo()
```